Formalizing Stack Safety as a Security Property\ifcameraready\else\vspace{-0.75cm}\fi

### 2023-05-23 submission5/core.tex
Polishing and comments

Thus, any low-level treatment of stack safety must begin by asking:
which high-level features are supported in a given
setting using the stack, and what are their
security requirements? Different answers will beget different properties.
\bcp{There's an awkward conceptual transition here: We pose a question
(``which high-level features are supported in a given
setting using the stack, and what are their
security requirements'') but then we don't seem to be going ahead and
answering it.  I suggest adding a sentence here that answers the
question directly, then a paragraph break.}
We identify and define five security requirements that a system might try
to enforce upon a function call: well-bracketed control flow, and the
integrity and confidentiality of each of the caller and the callee.
%
We apply these first to a simple system
with very few features, then to a more realistic one supporting
tail-call elimination, argument passing on the stack, and callee-save
registers.

We show that our properties are useful for distinguishing between correct and
incorrect enforcement using QuickChick~\cite{Denes:VSL2014,Pierce:SF4}, a property-based
random testing tool for Coq.
We find that {\LTC} is flawed in
a way that undermines both integrity and confidentiality; after
correcting this flaw, the repaired {\LTC} satisfies all our
properties.  Moreover, we modify {\LTC} to protect the features of
our more realistic system, and apply
random testing to validate this extended protection mechanism against
the extended properties.

@@

In sum, we offer the following contributions:

\begin{itemize}
\item We give a novel characterization of stack safety as a collection
  of properties: confidentiality and integrity for callee and caller,
  plus well-bracketed control-flow.
  The properties are parameterized over a notion of
  external observation, allowing them to characterize lazy enforcement
  mechanisms.
\item We extend these core definitions to
  describe a realistic setting with argument passing on the stack,
  callee-saves registers, and tail-call elimination. Our model is
  modular enough that handling these features is straightforward.
\item We validate a tag-based enforcement mechanism, \emph{Lazy
  Tagging and Clearing}, via property-based random testing, find that
it falls short, and propose and validate a fix.
\end{itemize}

In the next section, we give a brief overview of our framework and
assumptions. In \cref{sec:example}, we walk through a function call in
a simple example machine and discuss informally how each of our properties
applies to it. In the process we motivate the properties
from a security perspective.
%
In \cref{sec:formal} we formalize the machine model,
its {\em security semantics}, and the stack safety properties built on these.
\Cref{sec:extensions} describes how to support an extended set of features.
%\Cref{app:ptr} gives an (untested) proof-of-concept formalization of a more sophisticated,
%heap-like sharing model based on pointer provenance.
In \Cref{sec:enforcement} we describe the micro-policies that we test,
in \cref{sec:testing} the testing framework itself, and in
\cref{sec:relwork,sec:future} related and future work.

The accompanying artifact \apt{what is the state of that? Depending on conference preferences,
  should give a URL or other pointer.} \sna{Git repo is currently private, I think. I didn't see much
  in the way of conference preferences on a quick glance, but I'll look again in the morning.}
contains formal definitions (in Coq) of our
properties, plus our testing framework.  It does not include proofs:
we use Coq primarily for the QuickChick testing library and to
ensure that our definitions are unambiguous.  Formal proofs are left
as future work.

@@

Stack safety properties need to describe the behavior of machine code, but they naturally
talk about function activations and stack contents---abstractions that
are typically not visible at machine level. To bridge this gap,
our properties are defined in terms of a {\em security semantics} layered on top of
the standard execution semantics of the machine.  The security semantics identifies certain
state transitions of the machine as {\em security-relevant operations}, which update
a notional {\em security context}.  This context consists of
an (abstract) stack of function activations, each associated with a {\em view}
that maps each machine {\em state element} (memory location or register)
to a {\em security class} (active, sealed, etc.) specifying how the activation
can access the element.
The action of a security-relevant operation on the context is defined by a set
of rules that describe how the machine code implements the function abstractoin
using the stack and registers.

Given the security classes of the machine's state elements, we define high-level
security properties---integrity, confidentiality, and well-bracketed control flow---as
predicates that must hold on each call. These predicates draw on the idea of
\emph{variant} states from the theory of non-interference, and a notion of
{\em observable events}, which might include specific function calls (e.g., system
calls that perform I/O), writes to special addresses representing
memory-mapped regions, etc. For example, to show that certain locations are kept
secret, it suffices to compare the execution of states which vary at those locations
and check that their traces of observable events are the same. This structure
allows us to talk about the eventual implact of leaks or memory corruption without
reference to internal implementation details, and to support lazy enforcement by
flagging corruption of values only when it can actually impact visible behavior.

We introduce these properties by example in \cref{sec:example} and
formally in \cref{sec:formal}. In the remainder
of this section we describe the underlying framework in more detail.

\paragraph*{Machine Model}
We assume a conventional ISA (e.g. RISC-V, x86-64, etc.), with registers including a program counter
and stack pointer.
%, and a standard ABI.
%\sna{Not in our formalism (hardcoded +4 in the definition of WBCF notwithstanding.)
%  This assumption is made in our example and testing.}
We make no particular assumptions about the provenance of the machine code; in particular,
we do not assume the use of any particular compiler.
The machine may possibly be enhanced with % security
enforcement mechanisms such as hardware tags~\cite{pump_hasp2014,Gollapudi+23} or capabilities~\cite{Woodruff+14}. If so,
we assume that the behavior of these mechanisms is incorporated into the basic
step semantics of the machine, but we keep the enforcement state unchanged when constructing variant states.
Failstop behavior by enforcement mechanisms is modeled as stepping to the same state
(and thus silently diverging).
%% \sna{I think that the payload concept is too nitty-gritty for this section.
%%   The chief use of the payload is that it allows every value to be interpreted
%%   as an address. In the tag world, it happens to coincide with a natural instance
%%   of our variant-compatibility relation.}

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l | l | l |}
    \hline
    Operation \(\psi \in \Psi\) & Parameters & Sections\\
    \hline
    \example \(\mathbf{call}\) & target address, argument registers & \ref{sec:example},\ref{sec:formal}\\
    \testing & stack arguments (base, offset \& size) & \ref{sec:extensions},\ref{sec:testing} \\
    \example \(\mathbf{return}\) & & \ref{sec:example},\ref{sec:formal}\\
    \example \(\mathbf{alloc}\) & offset \& size & \ref{sec:example},\ref{sec:formal}\\
    \testing & public flag & \ref{sec:extensions},\ref{sec:testing} \\
    \example \(\mathbf{dealloc}\) & offset \& size &  \ref{sec:example},\ref{sec:formal}\\
    \testing \(\mathbf{tail call}\) & (same as for \(\mathbf{call}\)) & \ref{sec:extensions},\ref{sec:testing} \\
%    \hline
%    \multicolumn{2}{|c|}{{\it Pointer provenance operations}} \\
%    \hline
    \theory \(\mathbf{promote}\) & register, offset \& size & \ref{app:ptr} \\
    \theory \(\mathbf{propagate}\) & source register/address & \ref{app:ptr} \\
    \theory & destination register/address & \ref{app:ptr} \\
    \theory \(\mathbf{clear}\) & target register/address & \ref{app:ptr} \\
    \hline
  \end{tabular}
\end{center}
\caption{Security-relevant operations and their parameters, with the Sections in which
  they are first used in the paper.}
  \label{tab:psi}
\end{table}

@@

in a callee-save register (e.g. {\tt s0}) rather than in memory, its security class would still

@@

To begin with, what if {\tt f} returns to an unexpected place (i.e. \(\PCname \neq 20\) or
\(\SP \neq 980\))? We consider this to violate \(\wbcf\). \(\wbcf\) is a relationship between
call steps and their corresponding return steps: just after the return, the program
counter should be at the next instruction following the call,
and the stack pointer should be the same as it was before the call.
Both of these are essential for security. In \cref{subfig:WBCF}, the attacker adds
16 to the return address and then returns; this bypasses the {\tt if}-test in the code and outputs
{\tt secret}.
In \cref{subfig:WBCF2}, the attacker returns with \(\SP' = 988\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame,
\begin{center}
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  {\tt res} & {\tt sens} & {\tt sec} & \(\mbox{\tt ra}_1\) & \(\mbox{\tt ra}_2\) \\
  \hline
\end{tabular}
\end{center}

@@

A {\em machine state} \(\mach \in \MACHS\) is a map from state elements to a set \(\mathcal{V}\) of
\emph{values}.
Each value \(v\) contains a \emph{payload} word, written \(|v|\).
We write \(\mach[\component]\) to denote the value of \(\mach\) at
\(\component\)  and \(\mach[v]\) as shorthand for \(\mach[|v|]\).
Depending on the specific machine being modeled, values may also contain other
information relevant to hardware enforcement (such as a tag).
%Intuitively, the payload represents the part of the value that is relevant to
%the behavior of the basic machine, and it is the part that should be changed
When constructing variants (see~\cref{sec:props}, this additional information should
not be varied. To capture this idea, we assume a given \emph{compatibility} equivalence relation \(\sim\) on values,
and lift it element-wise to states.  Two values should be compatible if their
non-payload information (e.g. their tag) is identical.
%% \apt{This is still not very convincing!
%%   empting to say that only the payloads can affect the step function, but that is
%%   too strong, right? (OK for tags, but not, e.g. capabilities?}
%% \sna{Yeah, I don't think we can say something universal about when values should be
%%   considered compatible.}

@@

Lastly, what if we want to express a finer-grained notion of safety, in which
stack objects are protected unless the function that owns them intentionally
passes a pointer to them? This can be thought of as a {\it capability}-based
notion of security. Capabilities are unforgeable tokens that grant access to
a region of memory, typically corresponding to valid pointers to that region.
As such, this capability safety relies on some preexisting notion of pointer
validity, i.e. {\it pointer provenance}.
Memarian et al.'s PVI \cite{provenance} (provenance via integer)
memory model is a good option: it annotates pointers with the identity of the
object they first pointed to, and propagates the annotation when the
pointer is copied and when operations are performed on it.
This constitutes a substantial addition to the security context, which is why
this enhancement is more speculative than the others, and we have not tested it.

@@

%In particular, many enforcement techniques focus purely on
%well-bracketed control flow. For instance, stack canaries aim to prevent certain attacks on the return
%address, and shadow stacks with protection (e.g. Return Address Defender~\cite{Chiueh2001RAD})
%to enforce it completely. Others combine this protection with some degree of memory protection,
%chiefly focusing on integrity. Interestingly, Skorstengaard et al.~\cite{SkorstengaardSTKJFP}
%describe their ``local state encapsulation'' in terms of integrity, but it is equivalent
%to the combination of \(\clri\) and \(\clrc\). In fact, the follow-up work by
%Georges et al. \cite{Georges22:TempsDesCerises} was inspired by the properties presented in this paper to extend their formal discussion to include confidentiality.
%When checking if a preliminary version of these properties applied to
%the Cerise calling convention, they noted that the Skostengaard et al. calling convention
%did not enforce callee confidentiality, and made sure that their version would
%\cite{Georges22:personalcommunication}, in addition to building it into their formalism.
%This demonstrates the benefit of our choice to explicitly state properties in security
%terms: specifying security mechanisms is hard, and when the specification takes the
%form of a ``correct by construction'' machine, it is easy to neglect a non-obvious
%security requirement.

>>>>

Thus, any low-level treatment of stack safety must begin by asking:
which high-level features are supported in a given
setting using the stack, and what are their
security requirements? Different answers will beget different properties.
\bcp{There's an awkward conceptual transition here: We pose a question
(``which high-level features are supported in a given
setting using the stack, and what are their
security requirements'') but then we don't seem to be going ahead and
answering it.  I suggest adding a sentence here that answers the
question directly, then a paragraph break.} \bcp{Though, actually, the
next sentence repeats points that have already been made above.  Maybe
better to just delete it and find a new transition to the sentence
after.}
We identify and define five security requirements that a system might try
to enforce upon a function call: well-bracketed control flow, and the
integrity and confidentiality of each of the caller and the callee.
%
We apply these first to a simple system
with very few features, then to a more realistic one supporting
tail-call elimination, argument passing on the stack, and callee-save
registers.

We demonstrate the usefulness of our properties for distinguishing
between correct and
incorrect enforcement using
QuickChick~\cite{Denes:VSL2014,Pierce:SF4}, a property-based
random testing tool for Coq.
Indeed, we find that the published version of {\LTC} is flawed in
a way that undermines both integrity and confidentiality; after
correcting this flaw, {\LTC} satisfies all our
properties.  Further, we modify {\LTC} to protect the features of
our more realistic system and apply
random testing to validate this extended protection mechanism against
the extended properties.

@@

In summary, we offer the following contributions:

\begin{itemize}
\item We give a novel characterization of stack safety as a conjunction
  of security properties---confidentiality and integrity for callee
  and caller---plus well-bracketed control-flow.
  The properties are parameterized over a notion of
  external observation, allowing them to characterize lazy enforcement
  mechanisms.
\item We extend these core definitions to
  describe a realistic setting with argument passing on the stack,
  callee-saves registers, and tail-call elimination. The model is
  modular enough that adding these features is straightforward.
\item We validate a published enforcement mechanism, \emph{Lazy
  Tagging and Clearing}, via property-based random testing, find that
it falls short, and propose and validate a fix.
\end{itemize}
%
The following section offers a brief overview of our framework and
assumptions. \cref{sec:example} walks through a function call in
a simple example machine, discusses informally how each of our properties
applies to it, and motivates the properties
from a security perspective.
%
\cref{sec:formal} formalizes the machine model,
its {\em security semantics}, and the stack safety properties built on these.
\Cref{sec:extensions} describes how to support an extended set of features.
%\Cref{app:ptr} gives an (untested) proof-of-concept formalization of a more sophisticated,
%heap-like sharing model based on pointer provenance.
\Cref{sec:enforcement} describes the micro-policies that we test,
\cref{sec:testing} the testing framework itself, and
\cref{sec:relwork,sec:future} related and future work.

The accompanying artifact \apt{what is the state of that? Depending on conference preferences,
  should give a URL or other pointer.} \sna{Git repo is currently private, I think. I didn't see much
  in the way of conference preferences on a quick glance, but I'll look again in the morning.}
contains formal definitions (in Coq) of our
properties, plus our testing framework.  It does not include proofs, since
we use Coq primarily for the QuickChick testing library and to
ensure that our definitions are unambiguous.  Formal proofs are left
as future work.

@@

Stack safety properties need to describe the behavior of machine code, but they naturally
talk about function activations and stack contents---abstractions that
are typically not visible at machine level. To bridge this gap,
our properties are defined in terms of a {\em security semantics} layered on top of
the standard execution semantics of the machine.  The security semantics identifies certain
state transitions of the machine as {\em security-relevant operations}, which update
a notional {\em security context}.  This context consists of
an (abstract) stack of function activations, each associated with a {\em view}
that maps each machine {\em state element} (memory location or register)
to a {\em security class} (active, sealed, etc.) specifying how the activation
can access the element.
The action of a security-relevant operation on the context is defined by a set
of rules that characterize how particular machine code sequences are used
to implement the function abstraction in terms of the stack and registers.\bcp{OK?}

Given the security classes of the elements of the machine state, we
define high-level
security properties---integrity, confidentiality, and well-bracketed
control flow---as
predicates that must hold on each call. These predicates draw on the idea of
\emph{variant} states from the theory of non-interference, plus a notion of
{\em observable events}, which might include specific function calls (e.g., system
calls that perform I/O), writes to special addresses representing
memory-mapped regions, etc. For example, to show that certain locations are kept
secret, it suffices to compare executions starting at machine states which vary at those locations
and check that their traces of observable events are the same. This structure
allows us to talk about the eventual impact of leaks or memory corruption without
reference to internal implementation details and, in particular, to support lazy enforcement by
flagging corruption of values only when it can actually impact visible behavior.

We introduce these properties by example in \cref{sec:example} and
formally in \cref{sec:formal}. In the remainder
of this section we introduce the underlying \bcp{semantic?} framework
in more detail.

\paragraph*{Machine Model}
We assume a conventional ISA (e.g., RISC-V, x86-64, etc.), with registers including a program counter
and stack pointer.
%, and a standard ABI.
%\sna{Not in our formalism (hardcoded +4 in the definition of WBCF notwithstanding.)
%  This assumption is made in our example and testing.}
We make no particular assumptions about the provenance of the machine code; in particular,
we do not assume any particular compiler.
If the machine is enhanced with % security
enforcement mechanisms such as hardware
tags~\cite{pump_hasp2014,Gollapudi+23} or
capabilities~\cite{Woodruff+14}, we 
assume that the behavior of these mechanisms is incorporated into the basic
step semantics of the machine, but we keep the enforcement state
unchanged when constructing variant states\bcp{As a reader, I wonder
  ``How would they do that?  Seems like they are saying the tag
  enforcement or whatever is just part of the basic machine step
  relation and we can't see it, but then going on to place a
  requirement on it. How would such a requirement be phrased, formally?''}.
Failstop behavior by enforcement mechanisms is modeled as stepping to the same state
(and thus silently diverging).
%% \sna{I think that the payload concept is too nitty-gritty for this section.
%%   The chief use of the payload is that it allows every value to be interpreted
%%   as an address. In the tag world, it happens to coincide with a natural instance
%%   of our variant-compatibility relation.}

@@

\begin{table}
\begin{center}
  \begin{tabular}{| l | l | l |}
    \hline
    Operation \(\psi \in \Psi\) & Parameters & Sections\\
    \hline
    \example \(\mathbf{call}\) & target address, argument registers & \ref{sec:example},\ref{sec:formal}\\
    \testing & stack arguments (base, offset \& size) & \ref{sec:extensions},\ref{sec:testing} \\
    \example \(\mathbf{return}\) & & \ref{sec:example},\ref{sec:formal}\\
    \example \(\mathbf{alloc}\) & offset \& size & \ref{sec:example},\ref{sec:formal}\\
    \testing & public flag & \ref{sec:extensions},\ref{sec:testing} \\
    \example \(\mathbf{dealloc}\) & offset \& size &  \ref{sec:example},\ref{sec:formal}\\
    \testing \(\mathbf{tail call}\) & (same as for \(\mathbf{call}\)) & \ref{sec:extensions},\ref{sec:testing} \\
%    \hline
%    \multicolumn{2}{|c|}{{\it Pointer provenance operations}} \\
%    \hline
    \theory \(\mathbf{promote}\) & register, offset \& size & \ref{app:ptr} \\
    \theory \(\mathbf{propagate}\) & source register/address & \ref{app:ptr} \\
    \theory & destination register/address & \ref{app:ptr} \\
    \theory \(\mathbf{clear}\) & target register/address & \ref{app:ptr} \\
    \hline
  \end{tabular}
\end{center}
\caption{Security-relevant operations and their parameters, with the
  sections where
  they are first used.}
  \label{tab:psi}
\end{table}

@@

in a callee-save register (e.g., {\tt s0}) rather than in memory, its security class would still

@@

To begin with, what if {\tt f} returns to an unexpected place (i.e., \(\PCname \neq 20\) or
\(\SP \neq 980\))? We consider this to violate \(\wbcf\). \(\wbcf\) is a relationship between
call steps and their corresponding return steps: just after the return, the program
counter should be at the next instruction following the call,
and the stack pointer should be the same as it was before the call.
Both of these are essential for security. In \cref{subfig:WBCF}, the attacker adds
16 to the return address and then returns; this bypasses the {\tt if}-test in the code and outputs
{\tt secret}.
In \cref{subfig:WBCF2}, the attacker returns with \(\SP' = 988\) instead of the
correct \(\SP = 980\). In this scenario, given the layout of {\tt main}'s frame,
\begin{center}
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  {\tt res} & {\tt sens} & {\tt sec} & \(\mbox{\tt ra}_1\) & \(\mbox{\tt ra}_2\) \\
  \hline
\end{tabular}
\end{center}

@@

A {\em machine state} \(\mach \in \MACHS\) is a map from state elements to a set \(\mathcal{V}\) of
\emph{values}.
Each value \(v\) contains a \emph{payload} word, written \(|v|\).
We write \(\mach[\component]\) to denote the value of \(\mach\) at
\(\component\)  and \(\mach[v]\) as shorthand for \(\mach[|v|]\).
Depending on the specific machine being modeled, values may also contain other
information relevant to hardware enforcement (such as a tag).
%Intuitively, the payload represents the part of the value that is relevant to
%the behavior of the basic machine, and it is the part that should be changed
When constructing variants (see~\cref{sec:props}, this additional information should
not be varied. To capture this idea, we assume a given \emph{compatibility} equivalence relation \(\sim\) on values,
and lift it element-wise to states.  Two values should be compatible if their
non-payload information (e.g., their tag) is identical.
%% \apt{This is still not very convincing!
%%   empting to say that only the payloads can affect the step function, but that is
%%   too strong, right? (OK for tags, but not, e.g., capabilities?}
%% \sna{Yeah, I don't think we can say something universal about when values should be
%%   considered compatible.}

@@

Lastly, what if we want to express a finer-grained notion of safety, in which
stack objects are protected unless the function that owns them intentionally
passes a pointer to them? This can be thought of as a {\it capability}-based
notion of security. Capabilities are unforgeable tokens that grant access to
a region of memory, typically corresponding to valid pointers to that region.
As such, this capability safety relies on some preexisting notion of pointer
validity, i.e., {\it pointer provenance}.
Memarian et al.'s PVI \cite{provenance} (provenance via integer)
memory model is a good option: it annotates pointers with the identity of the
object they first pointed to, and propagates the annotation when the
pointer is copied and when operations are performed on it.
This constitutes a substantial addition to the security context, which is why
this enhancement is more speculative than the others, and we have not tested it.

@@

%In particular, many enforcement techniques focus purely on
%well-bracketed control flow. For instance, stack canaries aim to prevent certain attacks on the return
%address, and shadow stacks with protection (e.g., Return Address Defender~\cite{Chiueh2001RAD})
%to enforce it completely. Others combine this protection with some degree of memory protection,
%chiefly focusing on integrity. Interestingly, Skorstengaard et al.~\cite{SkorstengaardSTKJFP}
%describe their ``local state encapsulation'' in terms of integrity, but it is equivalent
%to the combination of \(\clri\) and \(\clrc\). In fact, the follow-up work by
%Georges et al. \cite{Georges22:TempsDesCerises} was inspired by the properties presented in this paper to extend their formal discussion to include confidentiality.
%When checking if a preliminary version of these properties applied to
%the Cerise calling convention, they noted that the Skostengaard et al. calling convention
%did not enforce callee confidentiality, and made sure that their version would
%\cite{Georges22:personalcommunication}, in addition to building it into their formalism.
%This demonstrates the benefit of our choice to explicitly state properties in security
%terms: specifying security mechanisms is hard, and when the specification takes the
%form of a ``correct by construction'' machine, it is easy to neglect a non-obvious
%security requirement.
